---
title: "all_childhooddementias"
author: "Karissa Barthelson"
date: "2024-12-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  autodep = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center", 
  out.width ="75%", 
  out.height = "75%"
)
```

```{r}
library(tidyverse)
library(GEOquery)
library(rentrez)
library(kableExtra)
library(readxl)
library(DT)
```

I want to write a script here to be able to recursively obtain all the dataset metadata in the GEO database for the childhood dementias. 

The tricky thing will be the keywords here, might use the childhood dementia knowledge-base as inspiration. 

## Key words
I went to the [CDI knowledgebase](https://knowledgebase.childhooddementia.org/) and downloaded a spreadsheet of "diseases". Below is a preview of this spreadsheet. 

```{r}
cdi_diseases <- read_excel("data/CDI_KB_diseases-1733292612.xlsx")

cdi_diseases %>%   
  head %>% 
  kable %>% 
  kable_styling() %>% 
  scroll_box(width = "100%", height = "800px")
```

The alternative_names column looks useful - carries all the names for each disease. 

I need to break this up into smaller fragments to be able to search this in R. rentrez will throw an error if I try searching all at once. 

```{r}
# split the multiple names within each disease
cd.keywords <- cdi_diseases$alternative_names %>% na.omit() %>% 
  str_split(pattern = ", ", ) %>% 
  unlist(use.names = F) %>% 
  str_split(pattern = "; ") %>% 
  unlist(use.names = F) %>% 
  unique
```

the cd keywords (`r length(cd.keywords)` words/terms) were exported for manual curation, then reimported. some of the abbreviations will not be totally useful and will pull out random stuff. 

```{r}
cd.keywords %>% 
  write.csv("output/cdi_keywords.csv")

cd.keywords <- 
  read.csv("output/cdi_keywords.csv", blank.lines.skip = TRUE)$x


# create a list which has 12 keywords per list element. 
cd.keywords.searchterm <- cd.keywords %>% 
  split(ceiling(seq_along(cd.keywords) / 5)) %>% # split the list
  lapply(function(x) {
    paste(x, "OR ") # add an OR for searching. 
  }) %>% 
  lapply(function(x) {
    x %>% 
      str_flatten() %>% # convert to a single vector
      str_sub(end = -5) # remove the last "OR"
  })

```

# all keywords. 
```{r}

# this takes a while
# search the GEO db for all the childhood dementia search terms. 
record.summaries.all <- cd.keywords.searchterm %>% 
  lapply(function(x) {
    search_results <- entrez_search(db = "gds", term = x)
    geo_accessions <- search_results$ids
    record_summaries <- lapply(geo_accessions, 
                               function(id) entrez_summary(db = "gds", id = id)
    )
})

# convert to a tibble
all.res.tibble <- record.summaries.all %>% 
  lapply(function(rs) {
    lapply(rs, function(x) 
      {
    tibble(
      accession = x$accession,    # GEO accession number
      species = x$taxon, 
      type = x$gdstype %>% unlist, # type of exp
      title = x$title,            # Title of the dataset
      summary = x$summary,         # Summary/description
      nsamples = x$samples %>% nrow
    )
    })
  }) %>% 
  bind_rows()
```

I then searched all the keywords recursively. The GEO database (GDS) was searched with the package rentrez for all the search terms from the "alternative names" column fron the CDI KB. This returned `r dim(all.res.tibble)[1]` datasets. These are found in the table below. you can search keywords in the search box. 

```{r}
# display as a DT table. 
all.res.tibble %>% 
  datatable(
    class = 'cell-border stripe', 
    filter = "top",
    caption = 'Table 1: childhood dementia datasets in GEO database.'
  )
```


## export data
A spreadsheet was generated for manual curation of irrelevant datasets. This can be downloaded from the github repo associated with this webpage. 

```{r}
# export the results for manual curation
record.summaries.all %>% 
  lapply(function(rs) {
    lapply(rs, function(x) 
      {
    tibble(
      accession = x$accession,    # GEO accession number
      species = x$taxon,
      type = x$gdstype %>% unlist, # type of exp
      title = x$title,            # Title of the dataset
      summary = x$summary,         # Summary/description
      nsamples = x$samples %>% nrow
    )
    })
  }) %>% 
  bind_rows() %>% 
  write_excel_csv("output/20241205 childhood dementia GEOdb.csv")
```


