---
title: "mucopolysacchidoses"
output:
  html_document:
    code_folding: hide
date: "2024-12-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  autodep = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center", 
  out.width ="75%", 
  out.height = "75%"
)
```

```{r}
library(tidyverse)
library(GEOquery)
library(rentrez)
library(kableExtra)
```

# Introduction

I want to write a script here to be able to recursively obtain all the dataset metadata in the GEO database for the childhood dementias. 

I will start as a proof of principal for mucopolysaccharidoses (MPS) diseases, as this has a nice keyword. 

```{r}
# Search GEO database for the term "mucopolysaccharidoses"
search_results <- entrez_search(db = "gds", term = "mucopolysaccharidoses")

# Get GEO accessions (GSE IDs)
geo_accessions <- search_results$ids


# get more info. 
# this takes a while
record_summaries <- lapply(geo_accessions, 
                           function(id) entrez_summary(db = "gds", id = id)
                           )
```

The GEO database (GDS) was searched with the package rentrez for the term "mucopolysaccharidoses". This returned `r search_results$count` datasets. 

Useful metadata has been added to the sheet below (and this can be exported - ask KB)

```{r}
lapply(record_summaries, function(x) {
  tibble(
    accession = x$accession,    # GEO accession number
    species = x$taxon,
    type = x$gdstype %>% unlist, # type of exp
    title = x$title,            # Title of the dataset
    summary = x$summary,         # Summary/description
    nsamples = x$samples %>% nrow
  )
}) %>% 
  bind_rows() %>% 
  kable(caption = "Datasets in GEO database containing mucopolysaccharidoses") %>% 
  kable_styling() %>% 
  scroll_box(width = "100%", height = "800px")
```

# conclusion 
This works nicely. Now to try expanding this to other disorders

